---
title: "Membrane cleaning MAG analysis"
editor: source
---

```{r}
library(tidyverse)
library(googlesheets4)
```

```{r}
user <- 
```

## Read drep (Widb) table

The drep table contains the information about the MAGs. The information that is most relevant to use is the completeness and contamination metrics.

```{r}
drep_table <- 
  '/scratch/users/snarayanasamy/membrane_cleaning/output/dereplication/all/dereplicated_bins/data_tables/Widb.csv' %>% 
  sprintf('ssh iris "cat %s"', .) %>% 
  pipe() %>% 
  read_csv() 
```

## Read CAT/BAT table with bin classification

This table contains the taxonomic classification of the bins

```{r}
bat_table <- 
  '/mnt/lscratch/users/snarayanasamy/membrane_cleaning/output/annotation/all/catbat/BAT.bin2classification.names_added.txt' %>% 
  sprintf('ssh iris "cat %s"', .) %>% 
  pipe() %>% 
  read_tsv() %>% 
  rename(genome = `# bin`)
```

Display the MAGs that have multiple assignments.

```{r}
bat_table %>% 
  filter(duplicated(genome)) %>% 
  select(genome, superkingdom, phylum, class, order, family, genus, species) %>% 
  knitr::kable()
```

As we can see, there are no support for most of the taxonomic levels, therefore, we exclude these bins from further analyses.

```{r}
bat_table <- 
  bat_table %>% 
  filter(!duplicated(genome)) 
```

```{r}
mag_info <- 
  inner_join(drep_table, bat_table, by = "genome") %>% 
  select(genome, completeness, contamination, size, N50, completeness_metric, contamination_metric, superkingdom, phylum, class, order, family, genus, species) %>% 
  separate(superkingdom, sep = ': ', into = c("superkingdom", "superkingdom_score")) %>% 
  separate(phylum, sep = ': ', into = c("phylum", "phylum_score")) %>% 
  separate(class, sep = ': ', into = c("class", "class_score")) %>% 
  separate(order, sep = ': ', into = c("order", "order_score")) %>% 
  separate(family, sep = ': ', into = c("family", "family_score")) %>% 
  separate(genus, sep = ': ', into = c("genus", "genus_score")) %>% 
  separate(species, sep = ': ', into = c("species", "species_score")) 
```

```{r}
taxa_of_interest <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1w5wCKSBxTwUp__m4_PcbmxbumGaY-9d90i75Gsb5DvQ/edit#gid=1591395475", sheet = "Taxa of interest")

inner_join(taxa_of_interest %>% select(genus), mag_info, by = "genus")
inner_join(taxa_of_interest %>% select(family), mag_info, by = "family")
inner_join(taxa_of_interest %>% select(order), mag_info, by = "order")
inner_join(taxa_of_interest %>% select(class), mag_info, by = "class")
inner_join(taxa_of_interest %>% select(phylum), mag_info, by = "phylum")

mag_info %>% filter(!is.na(species)) %>% filter(species != "no support")
```
