---
title: "Membrane cleaning MAG analysis"
author: "Shaman Narayanasamy"
editor: source
fig-cap-location: bottom
format:
  html: default
execute:
  echo: false 
  warning: false
  message: false
params:
  mag_info_ncbi: "/Users/naras0c/Work/data/membrane_cleaning/dereplicated_refined_MAG_information-ncbi.tsv"
  mag_info_gtdb: "/Users/naras0c/Work/data/membrane_cleaning/dereplicated_refined_MAG_information-gtdb.tsv"
---

```{r load_libraries}
library(tidyverse)
library(googlesheets4)
library(rlang)
library(phyloseq)
library(DESeq2)
library(vegan)

# Load the RColorBrewer library
library(RColorBrewer)
```
# Description

This document is generated from a quarto markdown file that contains the code for the analysis. To view the code, please visit the membrane cleaning github repository or request it from the author of this file. 

## Data
The following information is required for this analysis:

- `dRep` results, which includes completeness/contamination metrics from checkM
- `CAT/BAT` taxonomic classification output
- Gene annotations from `Bakta` to link transcripts to MAGs
- Mapping of Bakta genome IDs to MAG IDs 
- Raw counts of metagenomic reads mapped to MAGs
- Raw counts of metatranscriptomic reads mapped to transcripts of all MAGs
- Sample information table 

MAGs with multiple taxonomic assignments are considered to be ambiguous and are excluded from downstream analyses. The `dRep` table and `CAT/BAT` tables are merged to combine the completeness information with taxonomic information.
```{r read_mag_info}
mag_info <- 
  read_tsv(params$mag_info_ncbi)
```
## Expore the classification metrics
```{r visualise_mag_stats_pie_completeness}
#| label: fig-mag_stats_completeness
#| fig-cap: "Metagenomic assembled genome (MAG) completeness statistics." 
#| layout-ncol: 1
mag_info %>% 
  select("completeness_metric") %>% 
  group_by(completeness_metric) %>% 
  summarise(count = n()) %>% 
  ungroup() %>% 
 ggplot(aes(x = "", y = count, fill = completeness_metric)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  geom_text(aes(label = count), position = position_stack(vjust = 0.5)) +
  theme_void() +
  theme(legend.title = element_blank())
```
```{r visualise_mag_stats_pie_contamination}
#| label: fig-mag_stats_contamination
#| fig-cap: "Metagenomic assembled genome (MAG) contamination statistics." 
#| layout-ncol: 1
mag_info %>% 
  select("contamination_metric") %>% 
  group_by(contamination_metric) %>% 
  summarise(count = n()) %>% 
  ungroup() %>% 
 ggplot(aes(x = "", y = count, fill = contamination_metric)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  geom_text(aes(label = count), position = position_stack(vjust = 0.5)) +
  theme_void() +
  theme(legend.title = element_blank())
```
```{r create_ordering}
## Create taxonomic order for downstream visualisation
taxonomy_ordering <- c("superkingdom",
                       "phylum", 
                       "class", 
                       "order" ,
                       "family",
                       "genus",
                       "species"  
                       )
```

```{r visualise_mag_stats_bar}
mag_info %>% 
  select( 
    c("genome", 
      "superkingdom_score",
      "phylum_score",
      "class_score",
      "order_score" ,
      "family_score",
      "genus_score",
      "species_score" 
      )
  ) %>% 
  pivot_longer(cols = c(
      "superkingdom_score",
      "phylum_score",
      "class_score",
      "order_score" ,
      "family_score",
      "genus_score",
      "species_score"),
               names_to = "taxonomy", 
               values_to = "score"
               ) %>% 
  mutate(classified = ifelse(!is.na(score), "yes", "no")) %>% 
  separate(taxonomy, into = "taxonomy", sep = "_") %>% 
  group_by(taxonomy, classified) %>% 
  summarise(count = n()) %>% 
  ungroup() %>% 
  mutate(proportion = (count/257)*100,
         taxonomy = factor(taxonomy, levels = taxonomy_ordering)) %>% 
  ggplot(aes(x = taxonomy, y = proportion, fill = classified)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = c("yes" = "darkgreen", "no" = "gray")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, 
                                           margin = margin(t = 5, r = 0, b = 0, l = 0)))
```
## Explore the candidate bacteria of interest
```{r taxa_of_interest, eval = F}
taxa_of_interest <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1w5wCKSBxTwUp__m4_PcbmxbumGaY-9d90i75Gsb5DvQ/edit#gid=1591395475", sheet = "Taxa of interest")

inner_join(taxa_of_interest %>% select(genus), mag_info, by = "genus")
inner_join(taxa_of_interest %>% select(family), mag_info, by = "family")
inner_join(taxa_of_interest %>% select(order), mag_info, by = "order")
inner_join(taxa_of_interest %>% select(class), mag_info, by = "class")
inner_join(taxa_of_interest %>% select(phylum), mag_info, by = "phylum")

mag_info %>% filter(!is.na(species)) %>% filter(species != "no support")
```
```{r read_mag_info}
mag_info <- 
  mag_info %>% 
  ## Remove only those MAGs that were highly contaminated
  filter(contamination_metric != "very high") %>%  
  mutate_if(is.character, ~na_if(., "no support")) %>% 
  filter(!is.na(family) | genome == "TI2_MAGScoT_cleanbin_000096") ## Geobacter
```

```{r load_sample_information, warning=F}
sample_info <- 
  read_tsv('/Users/naras0c/Work/data/membrane_cleaning/sample_info.txt', comment = "#") %>% 
  dplyr::rename("sample_ID" = "sample_id")
```

## Metagenomics assembled genomes (MAG)
@fig-mag_stats shows the completeness statistics and the proportion of MAGs that
were successfully classified on each taxonomic level. Due to our stringent MAG
refinement and selection process, it is generally expected that we retain
relatively high-quality MAGs (@fig-mag_stats-1). However, only a small portion
of those MAGs could be classified at the lower taxonomy levels, e.g. class and
below (@fig-mag_stats-2). The larger number of "species" classified, is likely
due to the presence of MAGs from previous studies deposited into the database.


## Metagenomics 
@fig-mg_relab shows the metagenomic relative abundances of MAGs aggregated on
different taxonomic levels. The large number of unknown taxa on the lower
taxonomic levels (i.e. fig-mag_stats) is reflected in the respective
relative abundance plots (e.g. fig-mg_relab-4).

```{r prepare_mg_data}
mg_counts <- 
  read_tsv("/Users/naras0c/Work/data/membrane_cleaning/metagenomics_MAG_counts.tsv") %>% 
  pivot_longer(cols = 
                 sample_info %>% select(sample_ID) %>% 
                 separate(sample_ID, into = "sample_ID", sep = "_") %>% 
                 distinct() %>% pull(sample_ID),
               names_to = "sample_ID",
               values_to = "count") %>% 
  left_join(sample_info, by = "sample_ID") %>% 
  right_join(mag_info, by = c("MAG_ID" = "genome")) %>%  # Join based on the taxa that were retained for analysis
  mutate(phase = str_replace(phase, "normal", "initial"))
```


```{r write_mg_data, eval = F}
mg_counts %>% 
  pivot_wider(id_cols = "MAG_ID", names_from = "sample_ID", values_from = count) %>% 
  rowwise() %>% 
  mutate(mean = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  mutate(sd = sd(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(mag_info %>% select(MAG_ID = genome, 
                                superkingdom,
                                phylum,
                                order,
                                family,
                                genus,
                                species
                                ), by = "MAG_ID") %>%  
  select(MAG_ID,
         superkingdom, 
         phylum,
         order,
         family,
         genus,
         species, 
         everything()) %>% 
  arrange(desc(mean)) %>% 
  write_tsv("/Users/naras0c/Work/data/membrane_cleaning/metagenomics_MAG_counts_withMeanSD.tsv")

mg_counts %>% 
  select(sample_ID, MAG_ID, count) %>% 
  group_by(MAG_ID, sample_ID) %>% 
  mutate(count = sum(count)) %>% 
  ungroup() %>% 
  group_by(sample_ID) %>% 
  mutate(total_count_sample = sum(count)) %>% 
  mutate(relab = count/total_count_sample * 100) %>% 
  select(sample_ID, MAG_ID, relab) %>% 
  ungroup() %>% 
  pivot_wider(id_cols = "MAG_ID", names_from = "sample_ID", values_from = relab) %>% 
  rowwise() %>% 
  mutate(mean = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  mutate(sd = sd(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  ungroup() %>% 
  arrange(desc(mean)) %>% 
  left_join(mag_info %>% select(MAG_ID = genome, 
                                superkingdom,
                                phylum,
                                order,
                                family,
                                genus,
                                species
                                ), by = "MAG_ID") %>%  
  select(MAG_ID,
         superkingdom, 
         phylum,
         order,
         family,
         genus,
         species, 
         everything()) %>% 
  write_tsv("/Users/naras0c/Work/data/membrane_cleaning/metagenomics_MAG_relab_withMeanSD.tsv")
```

```{r create_palette, fig.width=12, fig.height=6}
# Create a palette with 33 colours and combine them
palette_33 <- c(brewer.pal(12, "Set3"),
                brewer.pal(8, "Set2"),
                brewer.pal(9, "Set1"),
                brewer.pal(4, "Dark2"))

# Verify the length of the palette
length(palette_33)  # Should be 33
```


```{r visualise_relab, fig.width=12, fig.height=6}
taxonomy_1 = "family"
taxonomy_2 = "genus"
  
taxonomy_sym_1 <- sym(taxonomy_1)
taxonomy_sym_2 <- sym(taxonomy_2)

mg_counts %>% 
  select(!!taxonomy_sym_1, !!taxonomy_sym_2, sample_ID, count, condition, phase, replicate) %>%
  mutate(taxonomy = paste(!!taxonomy_sym_1, !!taxonomy_sym_2, sep = ":")) %>% 
  mutate(phase = factor(phase, levels = c("initial", "backflush"))) %>% 

  group_by(taxonomy, sample_ID) %>% 
  mutate(count = sum(count)) %>% 
  ungroup() %>% 
  group_by(sample_ID) %>% 
  mutate(total_count_sample = sum(count)) %>% 
  mutate(relab = count/total_count_sample * 100) %>% 
  ungroup() %>% 
  mutate(taxonomy := ifelse(taxonomy == "no support", NA, taxonomy)) %>% 
  
  ggplot(aes(x = replicate, y = relab, fill = taxonomy) ) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = palette_33) +
  labs(
    y = "Relative abundance (%)",   
    x = "Cycle",
    fill = paste("Taxonomy", " - ", taxonomy_sym_1, ":", taxonomy_sym_2, sep = "")
  ) +
  facet_grid(phase ~ condition, scale = "free_x") +
  theme_bw() +
  theme(legend.position = "right", 
        strip.background = element_rect(color = NA) # Remove background
  )
```

## Metatranscriptomics
```{r load_sample_information, warning=F}
sample_info <- 
  read_tsv('/Users/naras0c/Work/data/membrane_cleaning/mt_sample_info.txt', comment = '#') %>% 
  mutate(sample_ID = paste(sample, lane, sep = '_'),
         lane = as.numeric(str_extract(lane, "\\d+"))) %>% 
  select(sample_ID, everything(), -sample, -R1, -R2) %>% 
  mutate(replicate = as.factor(replicate),
         lane = as.factor(lane)
         ) %>% 
  separate(sample_ID, extra = "drop", into = "sample", remove = FALSE) %>% 
  mutate(sample = str_remove_all(sample, "[0-9]$"))
```


```{r}
bakta2MAG <- read_tsv("/Users/naras0c/Work/data/membrane_cleaning/bakta2MAG_mappings.tsv")
```


```{r read_transcriptomics_data}
mt_counts <- 
  read_tsv("/Users/naras0c/Work/data/membrane_cleaning//metatranscriptomics_counts.tsv") %>% 
  select(Bakta_ID, contains("_L00")) %>% 
  pivot_longer(cols = sample_info %>% pull(sample_ID),
               names_to = "sample_ID",
               values_to = "count") %>% 
  group_by(Bakta_ID, sample_ID) %>% 
  summarise(count = sum(count, na.rm = T)) %>% 
  ungroup() %>% 
  distinct() %>% 
  left_join(bakta2MAG, by = "Bakta_ID") %>% 
  right_join(mag_info, by = c("MAG_ID" = "genome")) %>% 
  left_join(sample_info, by = "sample_ID")
```


```{r write_transcriptomics_data, eval = F}
mt_counts %>% 
  pivot_wider(id_cols = "MAG_ID", names_from = "sample_ID", values_from = count) %>% 
  ## Write out an intermediate file with the raw counts
  write_tsv("/Users/naras0c/Work/data/membrane_cleaning/metatranscriptomics_MAG_counts.tsv") %>% 
  rowwise() %>% 
  mutate(mean = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  mutate(sd = sd(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(mag_info %>% select(MAG_ID = genome, 
                                superkingdom,
                                phylum,
                                order,
                                family,
                                genus,
                                species
                                ), by = "MAG_ID") %>%  
  select(MAG_ID,
         superkingdom, 
         phylum,
         order,
         family,
         genus,
         species, 
         everything()) %>% 
  arrange(desc(mean)) %>% 
  ## Write file with mean and SD of the counts
  write_tsv("/Users/naras0c/Work/data/membrane_cleaning/metatranscriptomics_MAG_counts_withMeanSD.tsv")

mt_counts %>% 
  select(sample_ID, MAG_ID, count) %>% 
  group_by(MAG_ID, sample_ID) %>% 
  mutate(count = sum(count)) %>% 
  ungroup() %>% 
  group_by(sample_ID) %>% 
  mutate(total_count_sample = sum(count)) %>% 
  mutate(relab = count/total_count_sample * 100) %>% 
  select(sample_ID, MAG_ID, relab) %>% 
  ungroup() %>% 
  #mutate(relab_avg = sum(relab/23)) #%>% 
  pivot_wider(id_cols = "MAG_ID", names_from = "sample_ID", values_from = relab) %>% 
  rowwise() %>% 
  mutate(mean = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  mutate(sd = sd(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  ungroup() %>% 
  arrange(desc(mean)) %>%  
  left_join(mag_info %>% select(MAG_ID = genome, 
                                superkingdom,
                                phylum,
                                order,
                                family,
                                genus,
                                species
                                ), by = "MAG_ID") %>%  
  select(MAG_ID,
         superkingdom, 
         phylum,
         order,
         family,
         genus,
         species, 
         everything()) %>% 
  write_tsv("/Users/naras0c/Work/data/membrane_cleaning/metatranscriptomics_MAG_relab_withMeanSD.tsv")
```

```{r visualise_mt_relab, fig.width=12, fig.height=6}
taxonomy_1 = "family"
taxonomy_2 = "genus"
 
taxonomy_sym_1 <- sym(taxonomy_1)
taxonomy_sym_2 <- sym(taxonomy_2)

mt_counts %>% 
  rename("cycle" = "replicate") %>% 
  mutate(cycle_lane = paste(cycle, lane, sep = "_")) %>% 
  select(!!taxonomy_sym_1, !!taxonomy_sym_2, sample_ID, count, condition, phase, cycle, cycle_lane, lane) %>%
  mutate(taxonomy = paste(!!taxonomy_sym_1, !!taxonomy_sym_2, sep = ":")) %>% 
  mutate(phase = factor(phase, levels = c("initial", "backflush"))) %>% 

  group_by(taxonomy, sample_ID) %>% 
  mutate(count = sum(count)) %>% 
  ungroup() %>% 
  group_by(sample_ID) %>% 
  mutate(total_count_sample = sum(count)) %>% 
  mutate(relab = count/total_count_sample * 100) %>% 
  ungroup() %>% 
  mutate(taxonomy := ifelse(taxonomy == "no support", NA, taxonomy)) %>% 
  
  ggplot(aes(x = cycle_lane, y = relab, fill = taxonomy) ) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = palette_33) +
  labs(
    y = "Relative abundance (%)",   
    x = "Cycle_Lane",
    fill = paste("Taxonomy", " - ", taxonomy_sym_1, ":", taxonomy_sym_2, sep = "")
  ) +
  #facet_grid(~ phase + condition, scale = "free_x") +
  facet_grid(phase~condition, scale = "free_x") +
  theme_bw() +
  theme(legend.position = "right", 
        strip.background = element_rect(color = NA), # Remove background
        axis.text.x = element_text(angle = 90)
  )
```
```{r prepare_phyloseq}
## Phyloseq for MG data
mg_ps <- 
  phyloseq( 
      read_tsv("/Users/naras0c/Work/data/membrane_cleaning/metagenomics_MAG_counts.tsv") %>%  
        pivot_longer(cols = 
                       sample_info %>% 
                       select(sample_ID) %>% 
                       separate(sample_ID,  
                                into = c("sample_ID"), 
                                sep = "_",
                                remove = TRUE
                                ) %>% 
                       distinct() %>% 
                       pull(sample_ID),
               names_to = "sample_ID",
               values_to = "count") %>% 
      mutate(count = as.integer(count)) %>% 
      pivot_wider(id_cols = MAG_ID, names_from = sample_ID, values_from = count) %>% 
      column_to_rownames("MAG_ID") %>% 
      as.matrix() %>% 
      otu_table(taxa_are_rows = TRUE),
    sample_data( 
      sample_info %>%  
        select(-lane, -sample) %>% 
        separate(sample_ID,   
                 into = c("sample_ID"),  
                 sep = "_", 
                 remove = TRUE 
                 ) %>% 
        distinct() %>% 
        rename("cycle" = "replicate") %>%  
        #mutate(sample_ID = paste(sample_ID, cycle, sep = "")) %>% 
        #select(-sample) %>% 
        data.frame() %>% 
        distinct() %>% 
        column_to_rownames("sample_ID") %>% 
        sample_data()
    ), 
      read_tsv(params$mag_info_ncbi) %>% 
      select(MAG_ID = genome, c(taxonomy_ordering)) %>%  
      rename("kingdom" = "superkingdom") %>%  
      column_to_rownames("MAG_ID") %>%  
      as.matrix() %>% 
      tax_table()
)

mg_ps_rarefied <- rarefy_even_depth(mg_ps, 
                                rngseed = 1, # Set seed for reproducibility
                                
                                sample.size = 0.95*min(sample_sums(mg_ps)), # Subselect no. of reads equivalent to 95% of the smallest sample
                                replace = FALSE)
```

```{r remove_low_abundant_otus}
# define a minimum count threshold
min_count_threshold <- 1000000

# filter out otus with a total count below the threshold
mg_ps_pruned <- filter_taxa(mg_ps_rarefied, function(x) sum(x) >= min_count_threshold, TRUE)
```
```{r}
# Define a minimum sample presence threshold
min_samples_threshold <- round(nrow(sample_data(mg_ps_pruned)) * 0.10) # OTU must be present at least in 10% of the samples

# Filter out OTUs that are not present in at least 'min_samples_threshold' samples
mg_ps_pruned <- filter_taxa(mg_ps_pruned, function(x) sum(x > 0) >= min_samples_threshold, TRUE)

mg_ps_relab <- transform_sample_counts(mg_ps_pruned, function(x) x / sum(x))
```

### Alpha diversity
```{r compute_mg_richness}
plot_richness(mg_ps, color = "condition", measures = c("Observed", "Chao1"))

my_measures <- c("Observed", "Chao1", "Shannon", "Simpson")
plot_richness(mg_ps_pruned, x = "condition", measures = my_measures, nrow = 2) + 
  geom_boxplot(aes(fill = condition), alpha = 0.5) 

## Calculate richness based on difference measures
mg_rich <- estimate_richness(mg_ps_pruned)


# Test whether the observed number of OTUs differs significantly between conditions. Perform a non-parametric test, the Wilcoxon rank-sum test (Mann-Whitney):
pairwise.wilcox.test(mg_rich$Simpson, sample_data(mg_ps)$condition)
pairwise.wilcox.test(mg_rich$Shannon, sample_data(mg_ps)$condition)
pairwise.wilcox.test(mg_rich$InvSimpson, sample_data(mg_ps)$condition)
pairwise.wilcox.test(mg_rich$Fisher, sample_data(mg_ps)$condition)
```


```{r compute_mg_richness}
pairwise.wilcox.test(mg_rich$Simpson, sample_data(mg_ps)$phase)
pairwise.wilcox.test(mg_rich$Shannon, sample_data(mg_ps)$phase)
pairwise.wilcox.test(mg_rich$InvSimpson, sample_data(mg_ps)$phase)
pairwise.wilcox.test(mg_rich$Fisher, sample_data(mg_ps)$phase)
```


```{r compute_mg_richness}
pairwise.wilcox.test(mg_rich$Simpson, sample_data(mg_ps)$cycle)
pairwise.wilcox.test(mg_rich$Shannon, sample_data(mg_ps)$cycle)
pairwise.wilcox.test(mg_rich$InvSimpson, sample_data(mg_ps)$cycle)
pairwise.wilcox.test(mg_rich$Fisher, sample_data(mg_ps)$cycle)
```

### Beta diversity
```{r dimensionality_reduction, fig.width = 5, fig.height = 4}
# PCoA plot using the unweighted UniFrac as distance
mg_dist = phyloseq::distance(mg_ps_rarefied, method="bray", weighted=F)
ordination = ordinate(mg_ps_rarefied, method="PCoA", distance="bray")
plot_ordination(mg_ps_rarefied, ordination, color="condition") + theme(aspect.ratio=1)
```
```{r}
library(vegan)
adonis2(mg_dist ~ sample_data(mg_ps_rarefied)$cycle) 
adonis2(mg_dist ~ sample_data(mg_ps_rarefied)$condition)
adonis2(mg_dist ~ sample_data(mg_ps_rarefied)$phase)
```

## MT Diversity computation
```{r prepare_phyloseq}
## Phyloseq for MG data
mt_ps <- 
  phyloseq( 
  read_tsv("/Users/naras0c/Work/data/membrane_cleaning//metatranscriptomics_counts.tsv") %>% 
  select(Bakta_ID, contains("_L00")) %>% 
  pivot_longer(cols = sample_info %>% pull(sample_ID),
               names_to = "sample_ID",
               values_to = "count") %>% 
      mutate(count = as.integer(count)) %>% 
      pivot_wider(id_cols = MAG_ID, names_from = sample_ID, values_from = count) %>% 
      column_to_rownames("MAG_ID") %>% 
      as.matrix() %>% 
      otu_table(taxa_are_rows = TRUE),
    sample_data( 
      sample_info %>%  
        select(sample_ID, phase, condition, cycle = replicate) %>%  
        data.frame() %>% 
        distinct() %>% 
        column_to_rownames("sample_ID") %>% 
        sample_data()
    ), 
    mag_info %>% select(MAG_ID = genome, c(taxonomy_ordering)) %>%  
      rename("kingdom" = "superkingdom") %>%  
      column_to_rownames("MAG_ID") %>%  
      as.matrix() %>% 
      tax_table()
)

mt_ps_rarefied <- rarefy_even_depth(mt_ps, 
                                rngseed = 1, # Set seed for reproducibility
                                sample.size = 0.95*min(sample_sums(mt_ps)), # Subselect no. of reads equivalent to 95% of the smallest sample
                                replace = FALSE)
```

```{r remove_low_abundant_otus}
# define a minimum count threshold
min_count_threshold <- 1000000

# filter out otus with a total count below the threshold
mt_ps_pruned <- filter_taxa(mt_ps_rarefied, function(x) sum(x) >= min_count_threshold, TRUE)
```
```{r}
# Define a minimum sample presence threshold
min_samples_threshold <- round(nrow(sample_data(mt_ps_pruned)) * 0.10) # OTU must be present at least in 10% of the samples

# Filter out OTUs that are not present in at least 'min_samples_threshold' samples
mt_ps_pruned <- filter_taxa(mt_ps_pruned, function(x) sum(x > 0) >= min_samples_threshold, TRUE)

mt_ps_relab <- transform_sample_counts(mt_ps_pruned, function(x) x / sum(x))
```

### Alpha diversity
```{r compute_mt_richness}
plot_richness(mt_ps, color = "condition", measures = c("Observed", "Chao1"))

my_measures <- c("Observed", "Chao1", "Shannon", "Simpson")
plot_richness(mt_ps_pruned, x = "condition", measures = my_measures, nrow = 2) + 
  geom_boxplot(aes(fill = condition), alpha = 0.5) 

## Calculate richness based on difference measures
mt_rich <- estimate_richness(mt_ps_pruned)


# Test whether the observed number of OTUs differs significantly between conditions. Perform a non-parametric test, the Wilcoxon rank-sum test (Mann-Whitney):
pairwise.wilcox.test(mt_rich$Simpson, sample_data(mt_ps)$condition)
pairwise.wilcox.test(mt_rich$Shannon, sample_data(mt_ps)$condition)
pairwise.wilcox.test(mt_rich$InvSimpson, sample_data(mt_ps)$condition)
pairwise.wilcox.test(mt_rich$Fisher, sample_data(mt_ps)$condition)
```


```{r compute_mt_richness}
pairwise.wilcox.test(mt_rich$Simpson, sample_data(mt_ps)$phase)
pairwise.wilcox.test(mt_rich$Shannon, sample_data(mt_ps)$phase)
pairwise.wilcox.test(mt_rich$InvSimpson, sample_data(mt_ps)$phase)
pairwise.wilcox.test(mt_rich$Fisher, sample_data(mt_ps)$phase)
```


```{r compute_mt_richness}
pairwise.wilcox.test(mt_rich$Simpson, sample_data(mt_ps)$cycle)
pairwise.wilcox.test(mt_rich$Shannon, sample_data(mt_ps)$cycle)
pairwise.wilcox.test(mt_rich$InvSimpson, sample_data(mt_ps)$cycle)
pairwise.wilcox.test(mt_rich$Fisher, sample_data(mt_ps)$cycle)
```

### Dimensionality reduction
```{r dimensionality_reduction, fig.width = 5, fig.height = 4}
# PCoA plot using the unweighted UniFrac as distance
mt_dist = phyloseq::distance(mt_ps_rarefied, method="bray", weighted=F)
ordination = ordinate(mt_ps_rarefied, method="PCoA", distance="bray")
plot_ordination(mt_ps_rarefied, ordination, color="condition") + theme(aspect.ratio=1)
plot_ordination(mt_ps_rarefied, ordination, color="phase") + theme(aspect.ratio=1)
plot_ordination(mt_ps_rarefied, ordination, color="cycle") + theme(aspect.ratio=1)
```
```{r}
adonis2(mt_dist ~ sample_data(mt_ps_rarefied)$cycle) 
adonis2(mt_dist ~ sample_data(mt_ps_rarefied)$condition)
adonis2(mt_dist ~ sample_data(mt_ps_rarefied)$phase)
```
```{r deseq2}
## Define a model for downstream diffrential analysis. In this case, we are interested in the "condition" variable
sample_data(mg_ps)$condition <- as.factor(sample_data(mg_ps)$condition)

## Create differential abundance object for analysis object based on the model just created
mg_ds <- phyloseq_to_deseq2(mg_ps, ~ condition) 
mg_ds <- DESeq(mg_ds)

## Define the significance levels 
alpha = 0.01

## Compute the and rank different expression results that compares treatment (i.e. samples with OMP-spikes vs control)
mg_res_sig <- 
  results(mg_ds, contrast=list(c("condition_treatment_vs_control")), alpha=alpha) %>%  # Here is where we define the comparison
  as.data.frame() %>% 
  rownames_to_column("OTU") %>% 
  as_tibble() %>% 
  arrange(padj, desc(log2FoldChange)) %>% 
  filter(padj <= alpha)  %>%  
    left_join(  
      tax_table(mg_ps) %>% as.data.frame() %>% rownames_to_column("OTU") %>% as_tibble(),
      by = "OTU"
  ) %>% 
  filter(!is.na(genus)) # Remove any taxa unclassified on the Genus level
 
## Extract the top OTUs
top_otus <- mg_res_sig$OTU

## Display the table (for informational purposes)
select(mg_res_sig, # Remove the OTU column to clean up 
       c("kingdom", "phylum", "class", "order", "family", "genus", "species"), 
       everything(), 
       -OTU)
```

### Top OTUs/taxa
```{r get_top_abundances}
## From the differential abundance analyses, select the data of the most significant OTUs
top_otu_abundance <- 
  otu_table(ps, taxa_are_rows = TRUE) %>% t() %>%  
  as.data.frame() %>% rownames_to_column("OTU") %>% 
  as_tibble() %>% 
    filter(OTU %in% top_otus) %>% 
    pivot_longer(cols = -OTU, names_to = "sample_id", values_to = "abundance") %>% 
    left_join( 
      tax_table(ps)[top_otus,] %>% as.data.frame() %>% rownames_to_column("OTU") %>% as_tibble(), 
      by = "OTU") %>% 
    left_join( 
      sample_info,
      by = "sample_id"
    ) %>% select(-OTU) %>% 
  mutate(taxa = paste(Phylum, Genus, sep = ";")) %>% 
  select(-c(Kingdom, Phylum, Class, Order, Family, Genus, Species,)) %>% 
  group_by(taxa, sample_id) %>% 
  mutate(abundance = sum(abundance, na.rm = T)) %>% 
  ungroup() %>% 
  arrange(reactor, day) %>% 
  mutate(sample_name = paste(reactor, day, sep = ":")) %>% 
  mutate(condition = fct_relevel(condition, c("treatment", "control")),
         reactor = fct_relevel(reactor, c("A1", "A2", "B1", "B2")),
         day = as.character(day),
         day = fct_relevel(day, as.character(c(15, 30, 45, 60, 75, 90, 105, 120))),
         sample_name = fct_relevel(sample_name, c("A1:15", "A1:30", "A1:45",
         "A1:60", "A1:75", "A1:90", "A1:105","A1:120","A2:15", "A2:30", "A2:45",
         "A2:60", "A2:75", "A2:90", "A2:105","A2:120","B1:15", "B1:30", "B1:60",
         "B1:90", "B1:120","B2:15", "B2:30", "B2:60", "B2:90", "B2:120"))
         ) %>% 
  distinct()
```


```{r plot_heatmap}
top_otu_abundance %>% 
group_by(condition) %>% 
    tidyHeatmap::heatmap(.row = taxa, .column = sample_name, .value = abundance, scale = "column", 
                         cluster_rows = TRUE,  # Cluster rows to see if there are any intersting bacteria
                         cluster_columns = FALSE,  # Turn off clustering for columns
                         palette_value = circlize::colorRamp2(c(-2, -1, 0, 1, 2), viridis::magma(5))
                       ) %>% 
                         split_rows(5)
```


```{r plot_foldchange}
ggplot(res_sig, aes(x=Phylum, y=log2FoldChange, color=Genus)) +
    geom_jitter(size=3, width = 0.2) +
    theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
```
```{r}
## List top 10 taxa
res_sig %>% head(10) %>% 
  select(Phylum, Genus, padj, log2FoldChange) %>% 
  mutate(Taxa = paste(Phylum, Genus, sep = ";"),
         `log10(Fold change)` = log10(2^log2FoldChange),
         `-log10(Adjusted p-value)` = -log10(padj)) %>% 
  select(-c(Phylum, Genus, padj, log2FoldChange)) %>% 
  knitr::kable()
```

```{r deseq2}
## Define a model for downstream diffrential analysis. In this case, we are interested in the "condition" variable
sample_data(mt_ps)$condition <- as.factor(sample_data(mt_ps)$condition)

## Create differential abundance object for analysis object based on the model just created
mt_ds <- phyloseq_to_deseq2(mt_ps, ~ condition) 
mt_ds <- DESeq(mt_ds)

## Define the significance levels 
alpha = 0.01

## Compute the and rank different expression results that compares treatment (i.e. samples with OMP-spikes vs control)
mt_res_sig <- 
  results(mt_ds, contrast=list(c("condition_treatment_vs_control")), alpha=alpha) %>%  # Here is where we define the comparison
  
  as.data.frame() %>% 
  rownames_to_column("OTU") %>% 
  as_tibble() %>% 
  arrange(padj, desc(log2FoldChange)) %>% 
  filter(padj <= alpha)  %>%  
    left_join(  
      tax_table(mt_ps) %>% as.data.frame() %>% rownames_to_column("OTU") %>% as_tibble(),
      by = "OTU"
  ) %>% 
  filter(!is.na(genus)) # Remove any taxa unclassified on the Genus level
 
## Extract the top OTUs
top_otus <- mt_res_sig$OTU

## Display the table (for informational purposes)
select(mt_res_sig, # Remove the OTU column to clean up 
       "kingdom", "phylum", "class", "order", "family", "genus", "species", 
       everything(), 
       -OTU)

mt_res_sig %>% as.data.frame()
```